---
title: "Planned Restoration Data"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(leaflet)


knitr::opts_chunk$set(echo = TRUE)

nf_boundaries <- readRDS(here::here("data", "nf_boundaries.RDS"))
watersheds <- readRDS(here::here("data", "watersheds.RDS"))

```

Three datasets were provided by SIG:

-   [SOPA Forest Management](https://www.fs.usda.gov/sopa/index.php)

-   [CalVTP Proposed
    Projects](https://calfire-umb05.azurewebsites.net/projects-and-programs/calvtp-homepage-and-storymap/)

-   [USFS NEPA Planned
    Projects](https://data.fs.usda.gov/geodata/edw/datasets.php)

Due to size constraints, all raw data is stored here:
<https://drive.google.com/drive/u/0/folders/1EPi8wEZatXhZZfhvTjTcUZjuKSINT3o7>

**Objective**

Standardize these datasets into a single `planned_restoration` data
object with consistent formatting

### SOPA Forest Management

**Notes**

-   `Web_Link` is an invalid URL

```{r}
# USFS SOPA and CalVTP datasets
sopa_sf <- read_sf(here::here('data-raw', 'shapefiles', 'SOPA_Forest_Management_Scrape_Forest_HUC','SOPA_Forest_Management_Scrape_Forest_HUC.shp')) |>
  st_transform("+proj=longlat +datum=WGS84") |>
  glimpse()

table(sopa_sf$NFSLANDU_2)

sopa_sf_clean <- sopa_sf |> 
  select(project_name = Project_Na,
         national_forest_connection = NFSLANDU_2,
         watershed_name = Name,
         timing_plan = Decision, 
         contact = Project_co) |> 
  mutate(dataset = 'USFS SOPA') |> 
  glimpse()

```

### CALVTP Proposed Projects

**Challenges/Notes**

-   I have the dataset filtered to either falling within the HUC6
    watersheds OR the national forest boundaries. There are many more
    connections with the HUC6 than the forest boundaries so when
    integrated in the application, it will very few planned restorations
    for selected national forests. Most of the national forest column is
    `NA`.

```{r}
# CalVTP
cal_vtp_raw <- read_sf(here::here('data-raw', 'shapefiles', 'CalVTP_ProposedProjects_Pnt_Scrape_Forest_HUC1','CalVTP_ProposedProjects_Pnt_Scrape_Forest_HUC1.shp')) |>
  st_transform("+proj=longlat +datum=WGS84") |>
  glimpse()

table(cal_vtp_raw$CalVTP_P_3, useNA = 'always')

cal_vtp_clean <- cal_vtp_raw |> 
  select(project_name = L0CalVTP_P,
         national_forest_connection = CalVTP_P_3,
         watershed_name = CalVTP_P_8,
         timing_plan = L0CalVTP_8, 
         contact = L0CalVTP_5) |> 
  mutate(dataset = 'CalVTP') |> 
  glimpse()

```

### USFS NEPA Project Areas

**Notes**

-   Unsure what `rev_date` refers to but assuming that it is not the
    timing for the planned project. Unsure timing so leaving it as `NA`

-   To remain consistent with the other two datasets, took the centroid
    of the polygons

```{r}
# USFS
usfs_nepa <- read_sf(here::here('data-raw', 'shapefiles', 'USFS_NEPA_Project_Areas_3857_Forest_HUC','USFS_NEPA_Project_Areas_3857_Forest_HUC.shp')) |>
  st_transform("+proj=longlat +datum=WGS84") |>
  glimpse()

table(usfs_nepa$NFSLANDU_2, useNA = 'always')

usfs_nepa_clean <- usfs_nepa |> 
  select(project_name = name,
         national_forest_connection = NFSLANDU_2,
         watershed_name = name_1,
         contact = admin_di_1) |> 
  mutate(dataset = 'USFS EDW FACTS',
         timing_plan = NA) |> 
  filter(watershed_name %in% watersheds$name & national_forest_connection %in% nf_boundaries$name) |> 
  st_make_valid() |> 
  mutate(geometry = st_centroid(geometry)) |> 
  st_as_sf() |> 
  glimpse()
```

### Combine Datasets

**Notes**

-   Including all dataset records, number of rows: 1232 records

-   Filtering by national forest and huc6, number rows: 318

```{r}
planned_restoration <- bind_rows(cal_vtp_clean, 
                                 usfs_nepa_clean, # og: 685; filtered: 251
                                 sopa_sf_clean) |> 
  filter(watershed_name %in% watersheds$name & national_forest_connection %in% nf_boundaries$name) |> 
  glimpse()

saveRDS(planned_restoration, here::here("data", "planned_restoration.RDS"))

```

#### Map of all datasets

```{r}
pal_forest <- colorFactor(
  palette   = "Set3", 
  domain    = planned_restoration$national_forest_connection,
  na.color  = "lightgray"
)

leaflet(planned_restoration)  |> 
  addProviderTiles(providers$CartoDB.Positron)  |>    
  addCircleMarkers(
    radius       = 6,
    color        = ~pal_forest(national_forest_connection),
    stroke       = FALSE,
    fillOpacity  = 0.8,
    popup        = ~paste0(
      "<strong>Project Name:</strong> ", project_name, "<br/>",
      "<strong>National Forest:</strong> ", 
      ifelse(is.na(national_forest_connection), "None", national_forest_connection), "<br/>",
      "<strong>Watershed:</strong> ", watershed_name, "<br/>",
      "<strong>Timing Plan:</strong> ", timing_plan, "<br/>",
      "<strong>Contact:</strong> ", contact, "<br/>",
      "<strong>Dataset:</strong> ", dataset
    )
  )  |> 
  addLegend(
    position  = "bottomright",
    pal       = pal_forest,
    values    = ~national_forest_connection,
    title     = "National Forest"
  )

```
